---
title: "Activity: Bird Species"
author: "Mysha Nihal"
date: "06/07/2023"
output: html_notebook
---

# Set Up

```{r}
# Load `tidyverse` and `dcData` packages
library(tidyverse)
library(dcData)

# Load the `OrdwayBirds` data set into your RStudio environment from the `dcData` package

data("OrdwayBirds")

# Show a few commands used to inspect the data set a few different ways

head(OrdwayBirds)

tail(OrdwayBirds)

summary(OrdwayBirds)

```

# Activity

When you're finished with this activity, you will have a graph that shows what time of year various species appear at the Katherine Ordway Natural History Study Area in Inver Grove Heights, MN.

### Step 0

```{r}

# Get the data table & clean up dates (see Data Computing eBook)

Birds <- OrdwayBirds %>%
  select(SpeciesName, Month, Day) %>%
  mutate(Month = as.numeric(as.character(Month)),
         Day = as.numeric(as.character(Day)))

# Inspect resulting data

str(Birds)

```

### Step 1

```{r}
# unique species in original data (including mis-spellings)

SpeciesName <- unique(Birds$SpeciesName)

# unique species in the clean list

SpeciesName

```

### Step 2

```{r}

# inspect OrdwaySpeciesNames

data("OrdwaySpeciesNames")

# inner join

joinedData <- 
  inner_join(Birds, OrdwaySpeciesNames, by = "SpeciesName")

# inspect the data after the join

head(joinedData)
```

### Step 3

We want to identify the top 5-6 bird species sighted at the Katherine Ordway Natural History Study Area.

```{r}
# count sightings by species in descending order

species_count_des <- 
  joinedData %>% 
  count(SpeciesNameCleaned, sort = TRUE)

# a hint in the text book recommends displaying the top 10 species to choose our threshold

top_species <- head(species_count_des, 10)
top_species

# define a sightings threshold; produce a table with only birds belonging to major species

sightings_t <- 500
major_species <- species_count_des %>%
  filter(n >= sightings_t)
major_species

```

### Step 4

We now want to produce a month-by-month count for each of the major species, and then construct a chart to tell the story of what time of year the various species appear.

```{r}
# prepare glyph-ready data table called `ByMonth`

ByMonth <- joinedData %>%
  filter(SpeciesNameCleaned %in% major_species$SpeciesNameCleaned) %>%
  count(SpeciesNameCleaned, Month)

ByMonth

```

Now use the ggplot template below to generate the desired graph. (Remark you can also use esquisser or mplot to create a line chart instead of the barchart)

```{r}

ggplot(ByMonth) +
  aes(x = Month, y = n, colour = SpeciesNameCleaned) +
  geom_point(shape = "circle small", size = 1.5) +
  scale_color_hue(direction = 1) +
  theme_minimal()

```

According to the graph, we can answer the questions posed:

##### 1. Which species are present year-round?

- American Goldfinch

##### 2. Which species are migratory, that is, primarily present in one or two seasons?

- Robin

##### 3. What is the peak month for each major species? (bullet list)

-   Slate-colored Junco: 10
-   Tree Sparrow: 3 
-   Field Sparrow: 5 
-   American Goldfinch: 10 
-   Black-capped Chickadee: 11

##### 4. Which major species are seen in good numbers for at least 6 months of the year?

-   Blue Jay
-   Bluebird
-   Catbird
